Is deep learning good or bad for astrophysics?

### 2023-12-16 paper/main.tex
what the FUCK

% To do:
% ------
% - Write a title and abstract.
% - Write an outline.
% - Fill in the outline.
% - Publish on arXiv.
% - Format as ICML Position Paper (including title change).
% - Wait for the call from Stockholm.

@@

% Hogg typesetting issues
\setlength{\textwidth}{6.50in}
\setlength{\oddsidemargin}{0.00in}
\setlength{\textheight}{9.00in}
\setlength{\topmargin}{-0.50in}
\setlength{\parindent}{1.2\baselineskip} % seriously
\renewenvironment{abstract}{\paragraph{Abstract:}}{}
\linespread{1.08}
\frenchspacing\raggedbottom\sloppy\sloppypar
\renewcommand{\author}[1]{\medskip\par\noindent\textbf{#1}}
\newcommand{\affil}[1]{{\footnotesize\par\noindent #1 \par}}

@@

\section*{\raggedright %Position Paper:
Is machine learning good or bad for the natural sciences?}

@@

\begin{abstract}
  Machine learning (ML) methods are having a huge impact across all of the sciences.
  Advances have come from classification, regression, dimensionality reduction, clustering, emulation, and many other things.
  However, ML has a strong ontology---in which only the data exist---and a strong epistemology---in which a model is considered good if it performs well on held-out training data.
  These philosophies are in strong conflict with both standard practices and key philosophies in the natural sciences.
  Here I identify locations for ML in the natural sciences where the ontology and epistemology (and computational performance) are valuable.
  I also call out places in current practice where the introduction of ML introduces strong, unwanted statistical biases.
  For one, when ML models are used to emulate physical (or first-principles) simulations, they bring strong confirmation biases.
  For another, when flexible regressions are used to make datasets, the elements of those datasets cannot be used in joint analyses without taking on uncontrolled biases.
  On the other hand, there are also places where the introduction of ML can in fact make a project more scientifically conservative:
  When a flexible machine-learning model is used to represent the effects of nuisances, such as foregrounds, backgrounds, or instrument calibration parameters, the model capacity and loose philosophy of the method can sometimes make the results more trustworthy.
  These remarks are intended to apply to all of the natural sciences, but most of the specific examples are in the astrophysics domain.
\end{abstract}

@@

HOGG FIX: It is an understatement to say that deep learning is having a big impact across the sciences.
There are hundreds of papers per year now on deep-learning methods in astrophysics alone.
However, when we ask what scientific breakthroughs have been enabled by this influx of new tools and methods, there isn't a long list.
The success of the AlphaFold projects in protein structure (\cite{alphafold}) are often raised.
But these are successes in a biological or medical field, not astrophysics, and in a very specific challenge-problem setting in which performance is valued over understanding.
Here in astrophysics we care almost exclusively about understanding.

HOGG FIX: Astrophysics is an academic discipline, concerned with understanding the physical world, and naturally occurring mechanisms in play in that world.
We make progress by discovering new kinds of objects and phenomena, and explaining (and, even better, predicting) new kinds of objects and phenomena.
Our most successful investigations are judged in terms of the questions they answer, or the new questions they raise, or both.
The question here is: How can deep learning contribute to this mission?

HOGG FIX: I mentioned the word ``performance'' above.
One of the key ideas of machine-learning methods is that they are usually judged in terms of their performance, or their ability to predict or explain new data.
That's different, philosophically, from what's traditionally important in astrophysics.
When the expansion of the Universe was discovered (\cite{expansion, expansion2}), the result was not (primarily) judged on the basis of its ability to predict new data.
That is, the investigators were not mainly trying to predict the Doppler shifts of held-out galaxies!
They were trying to measure a property of the Universe, traced by galaxies.
And their result answered some important questions, and raised new ones.
Of course the expansion result has lasted in part because it \emph{does} make new predictions for new data, and very accurate predictions!
So it isn't irrelevant that the performance of the hypothesis is good.
However, there is some disconnect between how we assess a machine-learning method and how we assess a scientific contribution.

@@

@@

\section{Why do we need machine learning?}

>>>>

% To do:
% ------
% - Write an outline.
% - Fill in the outline.
% - Publish.
% - Wait for the call from Stockholm.

@@

% Hogg typesetting issues
\setlength{\textwidth}{6.50in}
\setlength{\oddsidemargin}{0.00in}
\setlength{\textheight}{9.00in}
\setlength{\topmargin}{-0.50in}
\setlength{\parindent}{1.2\baselineskip} % seriously
\linespread{1.08}
\frenchspacing\raggedbottom\sloppy\sloppypar
\renewcommand{\author}[1]{\medskip\par\noindent\textbf{#1}}
\newcommand{\affil}[1]{{\footnotesize\par\noindent #1 \par}}

@@

\section*{Is machine learning good or bad for the natural sciences?}

@@

\begin{abstract}
  Machine learning (ML) methods are having a huge impact across all of the sciences.
  Advances have come from classification, regression, dimensionality reduction, clustering, emulation, and many other things.
  However, ML has a strong ontology---in which only the data exist---and a strong epistemology---in which a model is considered good if it performs well on held-out training data.
  These philosophies are in strong conflict with both standard practices and key philosophies in the natural sciences.
  Here we identify locations for ML in the natural sciences where the ontology and epistemology (and computational performance) are valuable.
  We also show that the introduction of ML introduces strong, unwanted statistical biases.
  For one, when flexible models are used to emulate physical (or first-principles) simulations, they introduce strong confirmation biases.
  For another, when flexible regressions are used to make datasets, the elements of those datasets cannot be used in joint analyses without taking on uncontrolled biases.
  On the other hand, there are also places where the introduction of ML can in fact make a project more scientifically conservative; when a flexible machine-learning model is used to represent the effects of nuisances, such as foregrounds, backgrounds, or instrument calibration parameters, the model capacity and loose philosophy can sometimes make the results more trustworthy.
  These remarks are intended to apply to all of the natural sciences, but most of the specific examples are in the astrophysics domain.
\end{abstract}

@@

It is an understatement to say that deep learning is having a big impact across the sciences.
There are hundreds of papers per year now on deep-learning methods in astrophysics alone.
However, when we ask what scientific breakthroughs have been enabled by this influx of new tools and methods, there isn't a long list.
The success of the AlphaFold projects in protein structure (\cite{alphafold}) are often raised.
But these are successes in a biological or medical field, not astrophysics, and in a very specific challenge-problem setting in which performance is valued over understanding.
Here in astrophysics we care almost exclusively about understanding.

If we broaden our scope from deep learning to machine learning more generally, we can point to very important contributions of machine learning to astrophysics.
Kepler discoveries
PCA and, eg, SDSS redshift system

Astrophysics is an academic discipline, concerned with understanding the physical world, and naturally occurring mechanisms in play in that world.
We make progress by discovering new kinds of objects and phenomena, and explaining (and, even better, predicting) new kinds of objects and phenomena.
Our most successful investigations are judged in terms of the questions they answer, or the new questions they raise, or both.
The question here is: How can deep learning contribute to this mission?

I mentioned the word ``performance'' above.
One of the key ideas of machine-learning methods is that they are usually judged in terms of their performance, or their ability to predict or explain new data.
That's different, philosophically, from what's traditionally important in astrophysics.
When the expansion of the Universe was discovered (\cite{expansion, expansion2}), the result was not (primarily) judged on the basis of its ability to predict new data.
That is, the investigators were not mainly trying to predict the Doppler shifts of held-out galaxies!
They were trying to measure a property of the Universe, traced by galaxies.
And their result answered some important questions, and raised new ones.
Of course the expansion result has lasted in part because it \emph{does} make new predictions for new data, and very accurate predictions!
So it isn't irrelevant that the performance of the hypothesis is good.
However, there is some disconnect between how we assess a machine-learning method and how we assess a scientific contribution.

@@

I am both a developer of machine-learning methods, and a skeptic who thinks they should only rarely be used.
That said, there are reasons that we absolutely must use machine learning in the near future of astrophysics.
And there are indeed successes of machine learning that might seem invisible.
I'll try to speak to both of these points below.

@@

\section{Why do we need machine learning in the natural sciences?}
